---
title: "From Raw Ct to Insights: A qPCR Analysis Tutorial in R"
author: "[Your Name]"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5,
  dpi     = 150
)
```

---

## Overview

This tutorial walks through a complete qPCR analysis pipeline in R — from
simulated raw Ct values to statistical testing and publication-ready figures.

**What you will learn:**

- Simulate a realistic qPCR dataset
- Apply ΔCt and ΔΔCt normalization
- Run one-way and two-way ANOVA with Tukey HSD post-hoc
- Perform Principal Component Analysis (PCA) on expression data
- Build violin, boxplot, and interaction plots

**Experimental design used in this tutorial:**

| Factor | Levels |
|---|---|
| Group | Control, Treatment |
| Tissue | TypeA, TypeB |
| Replicates | 3 per condition |
| Gene panel | 20 target genes + 1 housekeeping |

All data are **simulated** — safe to adapt for any qPCR experiment.

---

## 1. Setup

```{r packages}
# Install any missing packages before loading
required <- c("tidyverse", "ggpubr", "RColorBrewer",
               "factoextra", "FactoMineR", "corrplot",
               "patchwork", "ggridges", "viridis", "gt")

to_install <- required[!required %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install)

library(tidyverse)
library(ggpubr)
library(RColorBrewer)
library(factoextra)
library(FactoMineR)
library(corrplot)
library(patchwork)
library(ggridges)
library(viridis)
library(gt)
```

---

## 2. Simulate qPCR Data

We simulate realistic Ct values for 20 target genes + 1 housekeeping gene
across two groups (Control / Treatment) and two tissue types (TypeA / TypeB),
with 3 technical replicates per sample.

```{r simulate-data}
set.seed(42)

# ── Experimental parameters ──────────────────────────────────────────────────
n_samples    <- 40        # total biological samples
n_targets    <- 20        # target genes
hk_gene      <- "GAPDH"  # housekeeping gene
groups       <- c("Control", "Treatment")
tissues      <- c("TypeA", "TypeB")
n_reps       <- 3         # technical replicates per sample

# ── Gene names ───────────────────────────────────────────────────────────────
target_genes <- paste0("Gene_", sprintf("%02d", 1:n_targets))
all_genes    <- c(target_genes, hk_gene)

# ── Sample metadata ──────────────────────────────────────────────────────────
metadata <- tibble(
  Sample_ID = paste0("S", sprintf("%02d", 1:n_samples)),
  Group     = rep(groups,  each = n_samples / 2),
  Tissue    = rep(tissues, times = n_samples / 2)
)

# ── Simulate raw Ct values ───────────────────────────────────────────────────
# Target genes: Ct ~ N(28, 2); Treatment lowers Ct by 1–3 for 8 "responsive" genes
# Housekeeping: Ct ~ N(18, 0.5) — stable across conditions (by design)

simulate_ct <- function(sample_meta, genes, hk) {
  map_dfr(1:nrow(sample_meta), function(i) {
    s <- sample_meta[i, ]

    # 8 genes respond to Treatment (lower Ct = higher expression)
    responsive  <- paste0("Gene_", sprintf("%02d", 1:8))
    effect      <- ifelse(s$Group == "Treatment" & genes %in% responsive,
                          runif(sum(genes %in% responsive), 1, 3), 0)

    # Tissue TypeB adds a small offset for genes 5–12
    tissue_eff  <- ifelse(s$Tissue == "TypeB" & genes %in% paste0("Gene_", sprintf("%02d", 5:12)),
                          runif(sum(genes %in% paste0("Gene_", sprintf("%02d", 5:12))), 0.5, 1.5), 0)

    ct_vals <- rnorm(length(genes), mean = 28, sd = 2) - effect - tissue_eff
    ct_hk   <- rnorm(1, mean = 18, sd = 0.5)

    tibble(
      Sample_ID = s$Sample_ID,
      Gene      = c(genes, hk),
      Ct        = c(ct_vals, ct_hk),
      Replicate = "mean"   # using mean Ct directly for clarity
    )
  })
}

ct_long <- simulate_ct(metadata, target_genes, hk_gene)
ct_wide <- ct_long %>%
  pivot_wider(names_from = Gene, values_from = Ct) %>%
  left_join(metadata, by = "Sample_ID")

cat("Simulated dataset:", nrow(metadata), "samples ×",
    length(all_genes), "genes\n")
glimpse(ct_wide %>% select(Sample_ID, Group, Tissue, Gene_01, Gene_02, GAPDH))
```

---

## 3. ΔCt Normalization

ΔCt subtracts the housekeeping gene Ct from each target gene Ct within the
same sample. This corrects for differences in RNA input and cDNA efficiency.

```{r delta-ct}
# ── Compute ΔCt for every target gene ────────────────────────────────────────
# ΔCt = Ct_target − Ct_housekeeping
# Lower ΔCt → higher relative expression

delta_ct <- ct_wide %>%
  mutate(across(all_of(target_genes),
                ~ . - GAPDH,
                .names = "dCt_{.col}")) %>%
  select(Sample_ID, Group, Tissue,
         starts_with("dCt_"))

# Rename for clarity: dCt_Gene_01 → Gene_01
colnames(delta_ct) <- gsub("^dCt_", "", colnames(delta_ct))

cat("ΔCt range across all genes and samples:\n")
delta_ct %>%
  pivot_longer(all_of(target_genes), names_to = "Gene", values_to = "dCt") %>%
  summarise(min = round(min(dCt), 2),
            mean = round(mean(dCt), 2),
            max = round(max(dCt), 2)) %>%
  print()
```

---

## 4. ΔΔCt and Fold Change

ΔΔCt compares each sample's ΔCt to the mean ΔCt of the reference group
(here: Control). Fold change = 2^(-ΔΔCt).

```{r ddct}
# ── Reference: mean ΔCt of Control group per gene ────────────────────────────
control_means <- delta_ct %>%
  filter(Group == "Control") %>%
  summarise(across(all_of(target_genes), mean)) %>%
  pivot_longer(everything(), names_to = "Gene", values_to = "ref_dCt")

# ── Compute ΔΔCt and fold change ─────────────────────────────────────────────
ddct_long <- delta_ct %>%
  pivot_longer(all_of(target_genes),
               names_to = "Gene", values_to = "dCt") %>%
  left_join(control_means, by = "Gene") %>%
  mutate(
    ddCt       = dCt - ref_dCt,              # ΔΔCt
    FoldChange = 2^(-ddCt)                   # relative expression
  )

# Summary: mean fold change per gene per group
fc_summary <- ddct_long %>%
  group_by(Gene, Group) %>%
  summarise(mean_FC = mean(FoldChange), .groups = "drop") %>%
  pivot_wider(names_from = Group, values_from = mean_FC) %>%
  mutate(FC_TreatvsControl = round(Treatment / Control, 2)) %>%
  arrange(desc(FC_TreatvsControl))

cat("Top 5 upregulated genes in Treatment vs Control:\n")
head(fc_summary, 5)
```

---

## 5. Descriptive Statistics

```{r descriptives}
# ── Summary stats: ΔCt by Group × Tissue ─────────────────────────────────────
dct_stats <- ddct_long %>%
  group_by(Gene, Group, Tissue) %>%
  summarise(
    n      = n(),
    mean   = round(mean(dCt),   2),
    sd     = round(sd(dCt),     2),
    se     = round(sd / sqrt(n), 3),
    median = round(median(dCt), 2),
    .groups = "drop"
  )

# Print a tidy table for the first 5 genes as a preview
dct_stats %>%
  filter(Gene %in% paste0("Gene_0", 1:5)) %>%
  gt() %>%
  tab_header(title    = "Descriptive statistics — ΔCt",
             subtitle = "First 5 genes, by Group × Tissue") %>%
  fmt_number(columns = c(mean, sd, se, median), decimals = 2) %>%
  tab_style(
    style = cell_fill(color = "#EAF3DE"),
    locations = cells_body(rows = Group == "Treatment")
  )
```

---

## 6. Statistical Testing

### 6.1 One-Way ANOVA: Group effect

```{r anova-oneway}
# ── One-way ANOVA: does expression differ by Group? ───────────────────────────
# Run for every gene; collect F-statistic and p-value

anova_group <- map_dfr(target_genes, function(g) {
  d   <- delta_ct %>% select(Group, Tissue, value = all_of(g))
  fit <- aov(value ~ Group, data = d)
  s   <- summary(fit)[[1]]
  tibble(Gene    = g,
         F_value = round(s$`F value`[1], 3),
         p_value = round(s$`Pr(>F)`[1],  4))
}) %>%
  mutate(q_value = round(p.adjust(p_value, method = "BH"), 4),
         Significant = q_value < 0.05) %>%
  arrange(p_value)

cat("Significant genes (BH q < 0.05):", sum(anova_group$Significant), "\n")
head(anova_group, 10)
```

### 6.2 Two-Way ANOVA: Group × Tissue interaction

```{r anova-twoway}
# ── Two-way ANOVA: Group × Tissue interaction ─────────────────────────────────
anova_interaction <- map_dfr(target_genes, function(g) {
  d   <- delta_ct %>% select(Group, Tissue, value = all_of(g))
  fit <- aov(value ~ Group * Tissue, data = d)
  s   <- summary(fit)[[1]]
  tibble(
    Gene              = g,
    F_Group           = round(s$`F value`[1], 3),
    p_Group           = round(s$`Pr(>F)`[1],  4),
    F_Tissue          = round(s$`F value`[2], 3),
    p_Tissue          = round(s$`Pr(>F)`[2],  4),
    F_Interaction     = round(s$`F value`[3], 3),
    p_Interaction     = round(s$`Pr(>F)`[3],  4)
  )
}) %>%
  arrange(p_Interaction)

cat("Genes with significant Group × Tissue interaction (p < 0.05):",
    sum(anova_interaction$p_Interaction < 0.05, na.rm = TRUE), "\n")
head(anova_interaction, 8)
```

### 6.3 Tukey HSD Post-Hoc

```{r tukey}
# ── Tukey HSD for the most significant gene ───────────────────────────────────
top_gene <- anova_group$Gene[1]
cat("Running Tukey HSD for:", top_gene, "\n\n")

d_top <- delta_ct %>%
  mutate(Group_Tissue = interaction(Group, Tissue)) %>%
  select(Group_Tissue, value = all_of(top_gene))

tukey_fit    <- aov(value ~ Group_Tissue, data = d_top)
tukey_result <- TukeyHSD(tukey_fit)
print(tukey_result)
```

---

## 7. Visualization

### 7.1 Mean ± SE Interaction Plot

```{r plot-interaction, fig.height=5}
# ── Pooled ΔCt summary across all genes ───────────────────────────────────────
pooled_summary <- ddct_long %>%
  group_by(Group, Tissue) %>%
  summarise(
    mean = mean(dCt),
    se   = sd(dCt) / sqrt(n()),
    .groups = "drop"
  )

ggplot(pooled_summary,
       aes(x = factor(Tissue, levels = c("TypeA", "TypeB")),
           y = mean, color = Group, group = Group)) +
  geom_point(size = 3.5) +
  geom_line(linewidth = 0.9) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.12) +
  scale_color_manual(values = c("Control" = "#1E8449", "Treatment" = "#C0392B")) +
  labs(x = "Tissue type", y = "Mean ΔCt (pooled across genes)",
       title = "Group × Tissue interaction",
       subtitle = "Mean ± SE; lower ΔCt = higher expression") +
  theme_classic(base_size = 13) +
  theme(axis.line  = element_line(linewidth = 1),
        axis.text  = element_text(face = "bold"),
        axis.title = element_text(face = "bold"),
        legend.title = element_text(face = "bold"))
```

### 7.2 Violin + Boxplot per Group

```{r plot-violin, fig.height=5}
# ── Long-format ΔCt for all genes ────────────────────────────────────────────
dct_long_plot <- delta_ct %>%
  pivot_longer(all_of(target_genes),
               names_to = "Gene", values_to = "dCt")

ggplot(dct_long_plot, aes(x = Tissue, y = dCt, fill = Group)) +
  geom_violin(alpha = 0.3, position = position_dodge(0.9), scale = "width") +
  geom_boxplot(width = 0.15, position = position_dodge(0.9),
               fill  = "white", outlier.size = 0.8) +
  stat_summary(fun = mean, geom = "point", size = 2,
               position = position_dodge(0.9), color = "black") +
  scale_fill_manual(values = c("Control" = "#1E8449", "Treatment" = "#C0392B")) +
  labs(x = "Tissue type", y = "ΔCt",
       title = "ΔCt distribution by Group and Tissue",
       subtitle = "Violin + boxplot + mean point") +
  theme_minimal(base_size = 13) +
  theme(axis.text  = element_text(face = "bold"),
        axis.title = element_text(face = "bold"),
        legend.title = element_text(face = "bold"))
```

### 7.3 Volcano-Style Fold Change Plot

```{r plot-volcano, fig.height=6}
# ── Gene-level fold change vs significance ────────────────────────────────────
volcano_df <- fc_summary %>%
  left_join(anova_group %>% select(Gene, q_value), by = "Gene") %>%
  mutate(
    log2FC    = log2(FC_TreatvsControl),
    neg_log10 = -log10(q_value),
    Status    = case_when(
      q_value < 0.05 & log2FC >  0.5 ~ "Up",
      q_value < 0.05 & log2FC < -0.5 ~ "Down",
      TRUE                            ~ "NS"
    )
  )

ggplot(volcano_df, aes(x = log2FC, y = neg_log10, color = Status)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed",
             color = "darkgreen", linewidth = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed",
             color = "red", linewidth = 0.6) +
  ggrepel::geom_text_repel(
    data = filter(volcano_df, Status != "NS"),
    aes(label = Gene), size = 3.5, max.overlaps = 15,
    fontface = "italic", family = "serif"
  ) +
  scale_color_manual(values = c("Up" = "#C0392B", "Down" = "#1565AC", "NS" = "grey60")) +
  labs(x     = expression(log[2]~"Fold Change (Treatment / Control)"),
       y     = expression(-log[10]~"(adjusted p-value)"),
       title = "Volcano plot — Treatment vs Control",
       color = "Expression") +
  theme_bw(base_size = 13)
```

### 7.4 Density Ridgeline Plot

```{r plot-ridgeline, fig.height=6}
# ── Ridgeline: ΔCt distribution per gene, coloured by Group ──────────────────
# Select the 10 most significant genes for readability
top10 <- anova_group$Gene[1:10]

dct_long_plot %>%
  filter(Gene %in% top10) %>%
  ggplot(aes(x = dCt, y = Gene, fill = Group)) +
  geom_density_ridges(alpha = 0.55, scale = 1.2,
                      rel_min_height = 0.01) +
  scale_fill_manual(values = c("Control" = "#1E8449", "Treatment" = "#C0392B")) +
  labs(x = "ΔCt", y = NULL,
       title = "ΔCt distribution — top 10 significant genes",
       subtitle = "Ridgeline density by Group") +
  theme_ridges(base_size = 12) +
  theme(legend.position = "top")
```

### 7.5 Heatmap-Style Tile Plot

```{r plot-heatmap, fig.height=6}
# ── Mean ΔCt heatmap: Gene × Group × Tissue ───────────────────────────────────
heatmap_df <- dct_long_plot %>%
  group_by(Gene, Group, Tissue) %>%
  summarise(mean_dCt = mean(dCt), .groups = "drop") %>%
  mutate(condition = paste0(Group, "\n", Tissue))

ggplot(heatmap_df, aes(x = condition, y = Gene, fill = mean_dCt)) +
  geom_tile(color = "white", linewidth = 0.4) +
  scale_fill_gradient2(
    low      = "#C0392B",
    mid      = "#F9F9F9",
    high     = "#1565AC",
    midpoint = median(heatmap_df$mean_dCt),
    name     = "Mean ΔCt"
  ) +
  labs(x = "Condition", y = "Gene",
       title = "Mean ΔCt heatmap",
       subtitle = "Red = high expression (low ΔCt), Blue = low expression") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(face = "bold"),
        panel.grid  = element_blank())
```

---

## 8. PCA of Gene Expression

PCA summarises the 20-gene expression profiles of each sample into a
low-dimensional space. Samples clustering by Group or Tissue confirm that
the gene panel captures relevant biological variation.

```{r pca, fig.height=5}
# ── PCA input: samples × genes (ΔCt matrix) ───────────────────────────────────
pca_matrix <- delta_ct %>%
  select(all_of(target_genes)) %>%
  as.data.frame()

rownames(pca_matrix) <- paste0(delta_ct$Sample_ID, "_",
                               delta_ct$Group, "_", delta_ct$Tissue)

pca_result <- PCA(pca_matrix, scale.unit = TRUE, ncp = 10, graph = FALSE)

# ── Scree plot ────────────────────────────────────────────────────────────────
p_scree <- fviz_eig(pca_result, addlabels = TRUE, ncp = 10,
                    ggtheme = theme_minimal(),
                    title   = "Scree plot — variance explained per PC")

# ── Sample map coloured by Group ──────────────────────────────────────────────
group_col <- delta_ct$Group

p_ind <- fviz_pca_ind(
  pca_result,
  geom.ind    = "point",
  col.ind     = group_col,
  palette     = c("Control" = "#1E8449", "Treatment" = "#C0392B"),
  addEllipses = TRUE,
  ellipse.type = "confidence",
  repel       = TRUE,
  legend.title = "Group",
  title        = "PCA — samples coloured by Group"
)

p_scree + p_ind
```

```{r pca-tissue, fig.height=5}
# ── Sample map coloured by Tissue ─────────────────────────────────────────────
tissue_col <- delta_ct$Tissue

p_tissue <- fviz_pca_ind(
  pca_result,
  geom.ind    = "point",
  col.ind     = tissue_col,
  palette     = c("TypeA" = "#1565AC", "TypeB" = "#E65100"),
  addEllipses = TRUE,
  ellipse.type = "confidence",
  repel        = TRUE,
  legend.title = "Tissue",
  title        = "PCA — samples coloured by Tissue"
)

# ── Variable (gene) map ───────────────────────────────────────────────────────
p_var <- fviz_pca_var(
  pca_result,
  col.var    = "contrib",
  gradient.cols = c("#74C476", "#005A32"),
  repel      = TRUE,
  title      = "PCA variable map — gene contributions"
)

p_tissue + p_var
```

```{r pca-contrib, fig.height=4}
# ── Top gene contributions to PC1 and PC2 ────────────────────────────────────
p_c1 <- fviz_contrib(pca_result, choice = "var", axes = 1, top = 10) +
  ggtitle("Gene contributions — PC1")
p_c2 <- fviz_contrib(pca_result, choice = "var", axes = 2, top = 10) +
  ggtitle("Gene contributions — PC2")

p_c1 + p_c2
```

---

## 9. Export Results

```{r export}
dir.create("results", showWarnings = FALSE)

# ΔCt normalized data
write.csv(delta_ct,       "results/delta_ct_normalized.csv",  row.names = FALSE)

# ΔΔCt and fold change
write.csv(ddct_long,      "results/ddct_fold_change.csv",     row.names = FALSE)

# ANOVA results
write.csv(anova_group,    "results/anova_group_effect.csv",   row.names = FALSE)
write.csv(anova_interaction, "results/anova_interaction.csv", row.names = FALSE)

# Fold change summary
write.csv(fc_summary,     "results/fold_change_summary.csv",  row.names = FALSE)

cat("Results exported to results/\n")
```

---

## 10. Session Info

```{r session}
sessionInfo()
```

---

## Key Formulas Reference

$$\Delta Ct = Ct_{target} - Ct_{housekeeping}$$

$$\Delta \Delta Ct = \Delta Ct_{sample} - \Delta Ct_{calibrator}$$

$$\text{Fold Change} = 2^{-\Delta\Delta Ct}$$

**Interpretation guide:**

| ΔCt | Expression level |
|---|---|
| Low (e.g. 2–5) | High expression |
| Medium (e.g. 8–12) | Moderate expression |
| High (e.g. >15) | Low / absent expression |

---

*This tutorial uses simulated data. Adapt the simulation parameters in
Section 2 to match your own experimental design.*
